% =========================================================
% APPENDIX A: Laplacian conditioning
% =========================================================
\chapter{Laplacian conditioning}
\label{app:laplacian}

\section{Residual training for Schr\"odinger}
\label{app:laplacian:setup}

We consider the stationary many-body Schr\"odinger equation on configuration space
$\mathbf R=(\mathbf r_1,\ldots,\mathbf r_N)\in\Omega\subset\mathbb R^{D}$ with $D=dN$
(for our quantum dots, $d=2$):
\begin{equation}
\hat H\Psi(\mathbf R)=E\Psi(\mathbf R),
\qquad
\hat H=\hat T+\hat V,
\qquad
\hat T=-\frac12\sum_{i=1}^N\Delta_i,
\label{eq:lap:sch}
\end{equation}
where $\hat V$ collects external and interaction potentials.

In \emph{strong residual minimization} one introduces
\begin{equation}
L := \hat H - E \quad\Rightarrow\quad L\Psi = 0,
\end{equation}
and minimizes a squared residual loss of the schematic form
\begin{equation}
\mathcal L_{\mathrm{res}}(\theta)
= \frac12\|L\Psi_\theta\|_{L^2(\Omega;w)}^2
= \frac12\int_\Omega |(L\Psi_\theta)(\mathbf R)|^2\,w(\mathbf R)\,d\mathbf R,
\label{eq:lap:resloss}
\end{equation}
for a nonnegative weight $w$ induced by the sampling strategy.
A key message of the numerical-analysis viewpoint is that strong residual losses
can be \emph{severely ill-conditioned}, and training speed is tightly linked to
spectral properties of an operator/matrix induced by $L$ \cite{De_Ryck_2024}.

\section{Linearized geometry and the matrix \(A\)}
\label{app:laplacian:A}

To expose the mechanism, follow the standard linearization argument used in
conditioning analyses \cite{De_Ryck_2024}.
Fix an expansion point $\theta_0$ and define tangent (feature) functions
\begin{equation}
q_j(\mathbf R) := \partial_{\theta_j}\Psi_\theta(\mathbf R)\big|_{\theta=\theta_0},
\qquad j=1,\ldots,P.
\end{equation}
For small parameter increments $\delta\theta$, we have the first-order model
\begin{equation}
\Psi_{\theta_0+\delta\theta}(\mathbf R)
\approx
\Psi_0(\mathbf R)+\sum_{j=1}^P q_j(\mathbf R)\,\delta\theta_j,
\qquad
\Psi_0:=\Psi_{\theta_0}.
\label{eq:lap:linPsi}
\end{equation}
Applying the residual operator gives
\begin{equation}
(L\Psi_{\theta_0+\delta\theta})(\mathbf R)
\approx
(L\Psi_0)(\mathbf R)+\sum_{j=1}^P (Lq_j)(\mathbf R)\,\delta\theta_j.
\label{eq:lap:linRes}
\end{equation}

Introduce the weighted inner product (complex-valued wavefunctions allowed)
\begin{equation}
\langle f,g\rangle_w := \int_\Omega f(\mathbf R)^\ast\,g(\mathbf R)\,w(\mathbf R)\,d\mathbf R.
\end{equation}
Let the Jacobian $J$ be the linear map whose $j$th column is $Lq_j$.
Then the loss \eqref{eq:lap:resloss} becomes (to second order in $\delta\theta$)
\begin{equation}
\mathcal L_{\mathrm{res}}(\theta_0+\delta\theta)
\approx
\frac12\|r_0 + J\,\delta\theta\|_w^2,
\qquad
r_0 := L\Psi_0.
\label{eq:lap:lsq}
\end{equation}
with normal-equation matrix
\begin{equation}
A := J^\ast J,
\qquad
A_{ij}=\langle Lq_i,\,Lq_j\rangle_w.
\label{eq:lap:Adef}
\end{equation}
Here $A$ is Hermitian positive semidefinite; in practice it can be close to singular,
so one often effectively works with a damped variant $A+\lambda I$ (implicit or explicit).

Now consider gradient descent on the quadratic model \eqref{eq:lap:lsq}
(with step size $\eta$). The error in parameter space evolves as
\begin{equation}
\delta\theta^{(t+1)}-\delta\theta^\star
=
\left(I-\eta A\right)\left(\delta\theta^{(t)}-\delta\theta^\star\right),
\label{eq:lap:GD}
\end{equation}
so each eigen-direction of $A$ contracts by a factor $(1-\eta\lambda)$.
Hence convergence speed is governed by the eigenvalue spread of $A$,
typically summarized by the (effective) condition number
\begin{equation}
\kappa(A)=\frac{\lambda_{\max}(A)}{\lambda_{\min}(A)}
\end{equation}
(over the nonzero/regularized spectrum).
Large $\kappa(A)$ implies stiffness: a step size small enough to be stable for
$\lambda_{\max}$ yields slow progress along directions associated with
$\lambda_{\min}$. This is precisely the conditioning bottleneck emphasized in
physics-informed training analyses \cite{De_Ryck_2024}.

\section{Bi-Laplacian scaling}
\label{app:laplacian:bilap}

The defining feature of Schr\"odinger is that $L$ contains a second-order
differential operator. To isolate the kinetic contribution, consider the regime
where the residual is dominated by the Laplacian part,
\begin{equation}
L \approx -\frac12\Delta_{\mathbf R}.
\label{eq:lap:Lapprox}
\end{equation}

\paragraph{From \(L\) to \(L^\ast L\).}
The matrix $A$ in \eqref{eq:lap:Adef} is a Gram matrix in the feature space
$\{Lq_j\}$.
The corresponding operator-level object is the \emph{Hermitian square}
$L^\ast L$ (with adjoint taken in the $L^2(\Omega;w)$ inner product)
\cite{De_Ryck_2024}. For general non-constant $w$, the adjoint $L^\ast$ differs from the
formal differential adjoint by weight-induced lower-order terms; the cleanest picture
is obtained for constant $w$ and standard boundary conditions (periodic/Dirichlet),
where the Laplacian is self-adjoint.

In the Laplacian-dominant regime \eqref{eq:lap:Lapprox}, one has (up to a constant factor)
\begin{equation}
L \approx -\tfrac12\Delta
\quad\Rightarrow\quad
L^\ast L \approx \tfrac14\,\Delta^2,
\label{eq:lap:bilap}
\end{equation}
i.e.\ a bi-Laplacian scaling. The prefactor does not affect conditioning ratios.

\paragraph{Fourier/eigenmode derivation of \(k^4\).}
On a periodic domain (or, more generally, in a Laplacian eigenbasis)
\(
-\Delta \phi_{\mathbf k} = |\mathbf k|^2\phi_{\mathbf k},
\)
so
\(
\Delta^2\phi_{\mathbf k} = |\mathbf k|^4\phi_{\mathbf k}.
\)
If an error $\delta\Psi$ expands as $\delta\Psi=\sum_{\mathbf k}c_{\mathbf k}\phi_{\mathbf k}$
and $L\approx -\tfrac12\Delta$, then
\begin{equation}
\|L\,\delta\Psi\|_{L^2}^2
\approx
\frac14\sum_{\mathbf k} |\mathbf k|^4|c_{\mathbf k}|^2.
\label{eq:lap:k4}
\end{equation}
Thus the squared residual penalizes high-frequency components with a \emph{quartic}
weight $|\mathbf k|^4$.
If the relevant spectrum spans $|\mathbf k|\in[k_{\min},k_{\max}]$, the induced
eigenvalue ratio scales like
\begin{equation}
\kappa \sim \left(\frac{k_{\max}}{k_{\min}}\right)^4,
\label{eq:lap:kappa}
\end{equation}
which is a sharp mathematical expression of ``Laplacian-driven stiffness''.
This quartic growth is consistent with the Laplacian/Poisson toy examples and
frequency-dependent conditioning effects emphasized in \cite{De_Ryck_2024}.

\paragraph{Length-scale interpretation.}
Let $\ell$ denote the smallest length scale that must be represented.
Then $k_{\max}\sim 1/\ell$ and \eqref{eq:lap:kappa} implies
\begin{equation}
\kappa \ \text{grows rapidly as}\ \ell\downarrow 0.
\end{equation}
In many-body Schr\"odinger problems, small $\ell$ arises from nodal structure,
strong localization, and short-range correlation features; in all such regimes,
the Laplacian makes the strong residual extremely curvature-sensitive.

\section{Barron/spectral perspective}
\label{app:laplacian:barron}

The conditioning story above can be rephrased in approximation terms.
In spectral Barron-type viewpoints, function classes are controlled by norms that
weight the Fourier transform by powers of frequency (or related spectral weights).
A key implication is that applying derivatives systematically increases the
importance of high-frequency tails \cite{De_Ryck_2024}.
Since $\Delta$ multiplies a Fourier mode by $|\mathbf k|^2$, and the squared
residual effectively involves $L^\ast L$ (hence $\Delta^2$ in Laplacian-dominant
regimes), the residual can require substantially more ``spectral budget'' than
the wavefunction itself.
In practical terms: even if $\Psi$ is approximable in a given network class,
\emph{approximating its strong residual} can be much harder, because the loss
implicitly demands accurate high-frequency curvature (cf.\ \eqref{eq:lap:k4}).
This is one of the motivations, in numerical analysis of PINNs, for considering
weak/variational formulations that reduce derivative order when the strong form
is too stiff \cite{De_Ryck_2024}.

\section{Implications for Stochastic Reconfiguration}
\label{app:laplacian:sr}

Stochastic reconfiguration (SR) updates parameters using a natural-gradient-like
preconditioner \cite{BeccaSorella2017-QMCBook,Amari1998-NaturalGradient,ParkKastoryano2020-GeometryNQS}.
Writing $O_i=\partial_{\theta_i}\log\Psi_\theta$, one common SR update is
\begin{equation}
S(\theta)\,\delta\theta = -g(\theta),
\qquad
S_{ij}=\mathrm{Cov}_{p_\theta}[O_i,O_j],
\qquad
g_i=\mathrm{Cov}_{p_\theta}[E_L,O_i],
\label{eq:lap:sr}
\end{equation}
where $p_\theta\propto|\Psi_\theta|^2$ and $E_L=(\hat H\Psi_\theta)/\Psi_\theta$.
SR improves conditioning in parameter space by using the (quantum) information
geometry of the variational family \cite{Amari1998-NaturalGradient,ParkKastoryano2020-GeometryNQS}.
However, the Laplacian-induced stiffness described above is an \emph{operator-level}
phenomenon: strong residuals magnify curvature errors via $\Delta$ and $\Delta^2$
(cf.\ \eqref{eq:lap:k4}), which remains a fundamental challenge even when parameter
updates are geometrically preconditioned.



% =========================================================
% APPENDIX B: Coulomb conditioning (2D, spin channels)
% =========================================================
\chapter{Coulomb conditioning}
\label{app:coulomb}

\section{Coulomb structure and singularity}
\label{app:coulomb:setup}

For our two-dimensional electronic systems ($\mathbf r_i\in\mathbb R^2$), the interaction
contains Coulomb terms of the form
\begin{equation}
\hat V_{\mathrm C}(\mathbf R)=\sum_{1\le i<j\le N}\frac{1}{r_{ij}},
\qquad r_{ij}=|\mathbf r_i-\mathbf r_j|.
\label{eq:coulomb:V}
\end{equation}
This is a \emph{singular multiplication operator}: it diverges at particle
coalescence $r_{ij}\to 0$.
The exact eigenfunction remains finite (or vanishes by antisymmetry) but develops
non-smooth short-distance structure (a cusp) so that divergences in kinetic and
potential contributions cancel in physically meaningful quantities such as the
local energy \cite{BeccaSorella2017-QMCBook}.

The central point for learning is:

\begin{quote}
Coulomb does not merely add ``another term'' to the residual; it introduces
\emph{singular short-range physics} that forces delicate cancellations with the
Laplacian. Strong residual objectives square these effects and become extremely
sensitive to tiny cusp errors.
\end{quote}

\section{2D cusp conditions: antiparallel vs parallel spins}
\label{app:coulomb:cusp}

We present the 2D cusp logic in a way that separates the two spin channels.
The cleanest statement (especially for practical ans\"atze) is in terms of the
short-distance behavior of a \emph{Slater--Jastrow} factorization
\begin{equation}
\Psi(\mathbf R)=D(\mathbf R)\,J(\mathbf R),
\qquad
J(\mathbf R)=\exp\!\Big(\sum_{i<j} u(r_{ij})\Big),
\label{eq:coulomb:slater-jastrow}
\end{equation}
because (i) the Slater part $D$ enforces antisymmetry and determines whether $\Psi$
vanishes at coalescence, and (ii) the Jastrow slope $u'(0)$ is what cancels the
$1/r$ Coulomb singularity in the local energy \cite{BeccaSorella2017-QMCBook}.

Fix a pair $(i,j)$ and introduce relative/center-of-mass coordinates in 2D:
\begin{equation}
\mathbf R_{ij} := \frac{\mathbf r_i+\mathbf r_j}{2},
\qquad
\mathbf r := \mathbf r_i-\mathbf r_j,
\qquad
r:=|\mathbf r|.
\end{equation}
A direct computation gives, in any dimension,
\begin{equation}
\Delta_i+\Delta_j = \frac12\,\Delta_{\mathbf R_{ij}} + 2\,\Delta_{\mathbf r},
\label{eq:coulomb:lap_decomp}
\end{equation}
hence the singular short-distance kinetic behavior is controlled by the relative term
$-\Delta_{\mathbf r}$.

A key 2D identity is
\begin{equation}
\Delta_{\mathbf r} r = \frac{1}{r}\qquad (r>0,\ \mathbf r\in\mathbb R^2).
\label{eq:coulomb:lap_r_2d}
\end{equation}

\paragraph{Antiparallel spins (distinguishable in the Slater part).}
If the two electrons have opposite spin, they enter different determinants in $D$
and generically $D$ does \emph{not} vanish at $\mathbf r\to 0$. Then $\Psi(0)\neq 0$ and
a direct cusp expansion in 2D is legitimate:
\begin{equation}
\Psi(\mathbf r,\ldots)
=
\Psi(0,\ldots)\Big(1 + a\,r + \mathcal O(r^2)\Big),
\qquad r\to 0.
\label{eq:coulomb:expansion_anti}
\end{equation}
Using \eqref{eq:coulomb:lap_r_2d}, the singular part of the relative Laplacian is
\begin{equation}
\Delta_{\mathbf r}\Psi \approx \Psi(0)\,a\,\Delta_{\mathbf r}r
= \Psi(0)\,\frac{a}{r},
\qquad
\frac{\Delta_{\mathbf r}\Psi}{\Psi}\approx \frac{a}{r}.
\label{eq:coulomb:lap_sing_2d}
\end{equation}
Since the pair kinetic contribution contains $-\Delta_{\mathbf r}$, the singular part
of the local energy is
\begin{equation}
E_L(\mathbf R)=\frac{\hat H\Psi}{\Psi}\supset \left(-\frac{a}{r}+\frac{1}{r}\right)
=\frac{1-a}{r}.
\label{eq:coulomb:EL_sing_2d_anti}
\end{equation}
Thus finiteness of $E_L$ at coalescence requires
\begin{equation}
a=1,
\qquad\text{i.e.}\qquad
\left.\partial_r \log \Psi\right|_{r=0}=1
\quad\text{(2D, antiparallel spins)}.
\label{eq:coulomb:cusp_2d_anti}
\end{equation}
Equivalently, in a Slater--Jastrow form with $D(0)\neq 0$,
this is the same as $\partial_r\log J|_{0}=1$ for the $ij$-pair.

\paragraph{Parallel spins (identical fermions, node at coalescence).}
For same-spin electrons, antisymmetry forces $D$ (and hence $\Psi$) to vanish as
$\mathbf r\to 0$. Generically one has a \emph{linear} zero:
\begin{equation}
D(\mathbf r,\ldots) \;=\; \bm\eta(\ldots)\cdot \mathbf r \;+\; \mathcal O(r^2),
\qquad r\to 0,
\label{eq:coulomb:D_linear_zero}
\end{equation}
with some nonzero vector $\bm\eta$ depending on the other coordinates.
In this case, the cancellation of the Coulomb singularity is controlled by the
Jastrow factor through cross terms in $-\Delta_{\mathbf r}(DJ)/(DJ)$.
A standard short-distance analysis (carried out explicitly for 2D quantum dots)
shows that the divergent kinetic contribution behaves like
\begin{equation}
-\frac{\Delta_{\mathbf r}(DJ)}{DJ}
\supset
-\frac{(2k+d-1)}{r}\,\partial_r\log J
\qquad\text{with}\qquad
d=2,\ \ k=1\ \text{(identical particles)},
\label{eq:coulomb:jastrow_div}
\end{equation}
while the Coulomb potential contributes $+1/r$ to $E_L$ as before.
Hence finiteness of the local energy requires
\begin{equation}
\left.\partial_r\log J\right|_{r=0}=\frac{1}{2k+d-1}=\frac{1}{3}
\quad\text{(2D, parallel spins)}.
\label{eq:coulomb:cusp_2d_par}
\end{equation}
This $1$ (antiparallel) versus $1/3$ (parallel) 2D cusp statement is the
standard quantum-dot/Jastrow--Slater cusp condition \cite{Siljamaki2003-QD-Cusps}.

\paragraph{What to remember.}
In 2D:
\begin{align}
\text{antiparallel spins:}\quad & \left.\partial_r\log \Psi\right|_{0}=1
\;\;\;\Longleftrightarrow\;\;\;
\left.\partial_r\log J\right|_{0}=1,
\label{eq:coulomb:summary_anti}
\\
\text{parallel spins:}\quad & \Psi(0)=0\ \text{(linear node)},\ \ 
\left.\partial_r\log J\right|_{0}=\frac13.
\label{eq:coulomb:summary_par}
\end{align}

\section{Why Coulomb is ill-conditioned (2D emphasis)}
\label{app:coulomb:residual}

The cusp conditions above show \emph{how} Coulomb creates stiffness: it forces
kinetic and potential contributions to be individually large and singular,
while their \emph{sum} is finite only under a short-distance constraint.
Strong residual minimization is therefore a \emph{cancellation problem}.

\paragraph{Sensitivity amplification in the local energy.}
If the learned antiparallel-spin cusp slope is $a_\theta = 1+\delta a$, then
\eqref{eq:coulomb:EL_sing_2d_anti} implies
\begin{equation}
E_L(\mathbf R)\supset -\frac{\delta a}{r},
\label{eq:coulomb:EL_mismatch_2d_anti}
\end{equation}
so an arbitrarily small cusp mismatch produces an arbitrarily large local-energy spike
as $r\to 0$.

For parallel spins, if the learned Jastrow slope is
$\partial_r\log J|_{0}=\frac13+\delta b$, then the same short-distance analysis leading to
\eqref{eq:coulomb:cusp_2d_par} implies
\begin{equation}
E_L(\mathbf R)\supset -\frac{3\,\delta b}{r}.
\label{eq:coulomb:EL_mismatch_2d_par}
\end{equation}
Thus both channels generate $1/r$ spikes unless the cusp is satisfied.

\paragraph{Why squaring is especially dangerous in 2D.}
A $1/r$ defect in $E_L$ (or in $(\hat H-E)\Psi$) is concentrated near coalescence.
But in 2D the squared singularity is only \emph{log-integrable}:
\[
\int_{r<\varepsilon}\Big(\frac{1}{r}\Big)^2\, d^2\mathbf r
=
\int_0^\varepsilon \frac{1}{r^2}\,(2\pi r\,dr)
=
2\pi\int_0^\varepsilon \frac{dr}{r}
=+\infty.
\]
Therefore, without effectively resolving the cusp, \emph{strong} squared-residual
objectives can be dominated (or even made ill-defined in the continuum limit) by the
coalescence region. In Monte Carlo practice, this manifests as heavy-tailed fluctuations
and extreme sensitivity to rare close-pair samples.

\paragraph{Residual expansion and cancellation structure.}
Writing $L=\hat T+(\hat V-E)$ and expanding the squared residual norm:
\begin{equation}
\|L\Psi\|_w^2
=
\|\hat T\Psi\|_w^2 + \|(\hat V-E)\Psi\|_w^2
+2\,\Re\langle \hat T\Psi,\;(\hat V-E)\Psi\rangle_w,
\label{eq:coulomb:expand}
\end{equation}
we see that near coalescence both $\hat T\Psi$ and $\hat V_{\mathrm C}\Psi$ are large.
The exact solution relies on cancellation between these contributions
(cf.\ \eqref{eq:coulomb:EL_sing_2d_anti} and \eqref{eq:coulomb:cusp_2d_par}), which means
\eqref{eq:coulomb:expand} demands high \emph{relative} accuracy in the cusp region:
if either term is off by a small fraction, the net residual can be large in absolute magnitude.
This is a classic recipe for ill-conditioning: the loss landscape contains directions
where small parameter changes cause huge residual changes in a small region.

\paragraph{Operator-level picture (\(L^\ast L\)).}
From the same viewpoint as Appendix~\ref{app:laplacian}, strong residual training
induces matrices/operators tied to $L^\ast L$ \cite{De_Ryck_2024}.
For $L=\hat T+(\hat V-E)$, one has (formally)
\begin{equation}
L^\ast L
=
\hat T^2 + (\hat V-E)^2 + \hat T(\hat V-E) + (\hat V-E)\hat T.
\label{eq:coulomb:LstarL}
\end{equation}
Coulomb enters here in two destabilizing ways:
\begin{enumerate}
\item the positive term $(\hat V-E)^2$ involves a $1/r^2$-type singular weight;
\item the mixed terms contain derivatives of $V$ through product rules, e.g.
$\hat T(V\Psi)$ generates $\nabla V\cdot\nabla\Psi$ and $\Delta V$ contributions,
which are even more singular/distributional at coalescence.
\end{enumerate}
This formal expansion explains why the Coulomb singularity can severely inflate
the effective spectral spread of the training operator, aligning with the
operator-conditioning viewpoint \cite{De_Ryck_2024}.

\section{Barron/spectral perspective}
\label{app:coulomb:barron}

Coulomb forces the \emph{true} 2D solution to have cusp-like, non-smooth short-range
structure, so approximating the \emph{strong residual} requires representing
high-frequency content that is not well captured by globally smooth/low-complexity
function classes.

In spectral/Barron-style approximation theories, approximation quality is governed
by spectral decay and norms that weight the Fourier tail \cite{De_Ryck_2024}.
A cusp (or a linear node times a cusp in $J$) implies slower spectral decay than a globally
smooth function.
Strong residual training then \emph{further} amplifies the high-frequency tail
because it applies $\Delta$ (and effectively $L^\ast L$), which multiplies Fourier
modes by powers of $|\mathbf k|$ (Appendix~\ref{app:laplacian}).
Hence, even if $\Psi$ can be approximated reasonably in function value, the strong
residual can remain large unless the cusp structure is represented accurately.
This is exactly the type of regularity mismatch that motivates weak/variational
formulations in numerical analysis of PINN-type methods, since weak forms reduce
the derivative order demanded of the approximation \cite{De_Ryck_2024}.

Practically, QMC wavefunctions therefore incorporate explicit short-range
correlation structure (e.g.\ cusp-enforcing Jastrow factors), precisely to remove
these singular residual pathologies and reduce the variance of local quantities
\cite{BeccaSorella2017-QMCBook,Siljamaki2003-QD-Cusps}.

\section{Implications for Stochastic Reconfiguration}
\label{app:coulomb:sr}

SR performs a natural-gradient step using a (quantum) Fisher/metric matrix
\cite{BeccaSorella2017-QMCBook,Amari1998-NaturalGradient,ParkKastoryano2020-GeometryNQS}.
In one common formulation,
\begin{equation}
S_{ij}=\mathrm{Cov}_{p_\theta}[O_i,O_j],
\qquad
g_i=\mathrm{Cov}_{p_\theta}[E_L,O_i],
\qquad
S\,\delta\theta=-g,
\label{eq:coulomb:sr}
\end{equation}
with $O_i=\partial_{\theta_i}\log\Psi_\theta$.

Coulomb-induced cusp mismatch produces $E_L$ spikes of the form
\eqref{eq:coulomb:EL_mismatch_2d_anti}--\eqref{eq:coulomb:EL_mismatch_2d_par},
which has two consequences:
\begin{enumerate}
\item \textbf{Heavy-tailed Monte Carlo forces.}
The estimator of $g_i=\mathrm{Cov}[E_L,O_i]$ becomes noisy when $E_L$ has rare but
extreme spikes (short-range events). This increases variance and can destabilize
the linear solve unless one uses damping/regularization, as standard in SR practice
\cite{BeccaSorella2017-QMCBook}.
\item \textbf{Coupling to curvature sensitivity.}
Even if SR corrects anisotropy in parameter space (natural gradient)
\cite{Amari1998-NaturalGradient,ParkKastoryano2020-GeometryNQS}, it does not remove
the underlying operator-driven stiffness: the spikes originate from imperfect
kinetic--potential cancellation enforced by Coulomb cusp physics.
\end{enumerate}

\paragraph{Bottom line.}
Laplacian dominance creates stiffness through $k^4$-type amplification of curvature
errors; Coulomb creates stiffness by enforcing cusp constraints (in 2D: $1$ vs $1/3$
depending on spin channel) and cancellation against the Laplacian.
Both mechanisms worsen conditioning of strong residual objectives and increase
statistical difficulty of SR through heavy-tailed local-energy fluctuations
\cite{De_Ryck_2024,BeccaSorella2017-QMCBook,Siljamaki2003-QD-Cusps}.

% ---------------------------------------------------------
% Add this BibTeX entry (new) if you want the 2D 1 vs 1/3 cusp citation:
% ---------------------------------------------------------
% @phdthesis{Siljamaki2003-QD-Cusps,
%   author = {Sami Siljam{\"a}ki},
%   title  = {Wave Function Methods for Quantum Dots in Magnetic Field},
%   school = {Helsinki University of Technology},
%   year   = {2003},
%   note   = {See Sec.~2.4 (Cusp conditions), Eq.~(2.27) and the 2D specialization giving
%             $\partial_r J/J = C$ (antiparallel) and $\partial_r J/J = C/3$ (parallel).}
% }


% =========================================================
% APPENDIX C: The Collocation--Backflow Catch-22
% =========================================================
\chapter{The collocation--backflow catch-22}
\label{app:catch22}

This appendix formalizes a structural difficulty encountered when training
backflow transformations under collocation (strong residual) objectives
and presents extensive numerical evidence for the $N{=}6$, $\omega{=}0.5$
quantum dot.
The core observation is a \emph{catch-22}: backflow's learning signal is
concentrated near the nodal surface, but
the local energy diverges there for any imperfect wavefunction, and any
sampling strategy that includes these points produces
gradients with unbounded variance.
The result is an irreducible accuracy floor that is absent
under VMC\,+\,SR training of the same ansatz.

\section{Setup and notation}
\label{app:catch22:setup}

We consider a Slater--Jastrow--backflow ans\"atz of the form
\begin{equation}
\Psi_\theta(\mathbf R)
=\det\tilde S_\uparrow(\mathbf R;\theta_{\mathrm{BF}})
 \;\det\tilde S_\downarrow(\mathbf R;\theta_{\mathrm{BF}})
 \;\exp\bigl(J(\mathbf R;\theta_J)\bigr),
\label{eq:c22:ansatz}
\end{equation}
where $\tilde S_\sigma$ is the Slater matrix evaluated at backflow-shifted
coordinates $\tilde{\mathbf r}_i = \mathbf r_i + \Delta\mathbf
x_i(\mathbf R;\theta_{\mathrm{BF}})$, $J$ is a neural Jastrow factor
(PINN), and $\theta=(\theta_J,\theta_{\mathrm{BF}})$ are trainable
parameters.
A CTNN message-passing network produces the backflow shifts $\Delta\mathbf x_i$.

We train by minimizing the \emph{screened collocation} objective
\begin{equation}
\mathcal L(\theta)
=\operatorname{Var}_{\mathcal S}\!\bigl[E_L(\mathbf R;\theta)\bigr],
\qquad
E_L=\frac{\hat H\Psi_\theta}{\Psi_\theta},
\label{eq:c22:loss}
\end{equation}
where the collocation set $\mathcal S$ is obtained by
importance-sampling from a Gaussian proposal~$q$
and retaining the top-$K$ points ranked by $|\Psi_\theta|^2/q$
(``screened collocation'').
This top-$K$ selection naturally suppresses near-node configurations
where $|\Psi|^2\approx 0$.


\section{The divergence of $E_L$ near nodes}
\label{app:catch22:divergence}

The local energy decomposes as
\begin{equation}
E_L
= -\frac12\frac{\nabla^2\Psi}{\Psi}+V
= -\frac12\Bigl[
  \nabla^2 J + |\nabla J|^2
  + 2\,\nabla J\cdot\frac{\nabla D}{D}
  + \frac{\nabla^2 D}{D}
\Bigr]+V,
\label{eq:c22:EL_decomp}
\end{equation}
where $D:=\det\tilde S_\uparrow\det\tilde S_\downarrow$ and
the last term $\nabla^2 D/D$ contains second derivatives of the
backflow shifts $\nabla^2\Delta\mathbf x_i$ amplified by the
inverse determinant.

\paragraph{For the exact wavefunction, $E_L$ is constant.}
If $\Psi$ is an eigenstate, $E_L=E$ everywhere, and the apparent
$D^{-1}$ singularity is cancelled by the numerator vanishing at the same
rate as the denominator.
Concretely, for a smooth wavefunction with a simple node at surface
$\mathcal N$, both $D$ and $\nabla^2 D$ vanish linearly in the
signed distance~$d$ from~$\mathcal N$:
\begin{equation}
D(\mathbf R) = \mathcal O(d),
\qquad
\nabla^2 D(\mathbf R) = \mathcal O(d),
\qquad
\frac{\nabla^2 D}{D}\;\text{is finite on }\mathcal N.
\label{eq:c22:exact_cancel}
\end{equation}

\paragraph{For an approximate wavefunction, the cancellation generically fails.}
During training, the backflow parameters $\theta_{\mathrm{BF}}$ are
imperfect: the \emph{learned} nodal surface $\mathcal N_\theta$ (where
$D(\cdot;\theta)=0$) does not coincide with the \emph{true} nodal surface.
Let $d_\theta$ denote the signed distance from the learned node and $d^*$
that from the true node.
Near a point where $d_\theta\approx 0$ but $d^*\neq 0$, the determinant
$D\sim d_\theta\to 0$ while the kinetic contributions from the Jastrow and
potential terms remain finite:
\begin{equation}
E_L(\mathbf R)
\supset
\frac{\nabla^2 D(\mathbf R;\theta)}{D(\mathbf R;\theta)}
\sim \frac{c}{d_\theta}\to\pm\infty
\qquad\text{as }d_\theta\to 0,\ d^*\neq 0,
\label{eq:c22:EL_div}
\end{equation}
for some configuration-dependent constant $c\neq 0$.
Conversely, near a point on the true node ($d^*=0$) but off the learned node
($d_\theta\neq 0$), $\Psi\sim d^*\to 0$ and the wavefunction assigns
vanishing weight.
The pathology \eqref{eq:c22:EL_div} is therefore not an artifact of the
nodal structure itself---it is a property of \emph{mismatched} nodes,
and is worst precisely when backflow still has something to learn.


\section{The geometric mismatch argument}
\label{app:catch22:mismatch}

We now show that the variance of the local energy (and hence of the
training gradients) generically diverges for imperfect backflow parameters.

Consider the contribution to $\operatorname{Var}[E_L]$ from a
neighborhood $U_\varepsilon$ of the learned nodal surface:
\begin{equation}
\operatorname{Var}[E_L]
\ge
\int_{U_\varepsilon}
|\Psi(\mathbf R;\theta)|^2\,
\bigl(E_L(\mathbf R;\theta)-\bar E\bigr)^2\,
d\mathbf R.
\label{eq:c22:var_lb}
\end{equation}
When the node is ``simple'' ($D$ vanishes linearly), we have
\begin{align}
|\Psi|^2 &\sim d_\theta^2, &
E_L^2 &\sim d_\theta^{-2\alpha},
\label{eq:c22:scaling}
\end{align}
where $\alpha$ characterises the severity of the mismatch.
For the generic case where the numerator
$\nabla^2 D$ does \emph{not} vanish on $\mathcal N_\theta$
(because $\mathcal N_\theta\neq\mathcal N^*$),
one has $\alpha=1$ so that $E_L^2\sim d_\theta^{-2}$.

In $D$-dimensional configuration space, the volume element in a
tubular neighborhood of a codimension-1 surface scales as
$dV\sim d_\theta^0\,d\sigma\,dd_\theta$.
Thus the integrand in \eqref{eq:c22:var_lb} behaves as
\begin{equation}
|\Psi|^2\cdot E_L^2\cdot dV
\sim
d_\theta^2\cdot d_\theta^{-2}\cdot dd_\theta
= dd_\theta,
\label{eq:c22:integrand}
\end{equation}
which is just integrable (logarithmically).
But the key observation is that this is the \emph{best case}.
In practice, the mismatch between $\mathcal N_\theta$ and $\mathcal N^*$
is not uniform: there exist regions where the surfaces cross at angles,
creating pockets where $|d^*|/|d_\theta|$ is large
(the true wavefunction is non-negligible while the learned determinant
is near-zero), and there the $E_L^2$ spike is not compensated by $|\Psi|^2$
suppression.

More concretely, the gradient
$\partial\mathcal L/\partial\theta_{\mathrm{BF}}$ involves terms of the form
\begin{equation}
\frac{\partial E_L}{\partial\theta_{\mathrm{BF}}}
\;\supset\;
\frac{1}{D}\frac{\partial(\nabla^2 D)}{\partial\theta_{\mathrm{BF}}}
\;-\;
\frac{\nabla^2 D}{D^2}
\frac{\partial D}{\partial\theta_{\mathrm{BF}}},
\label{eq:c22:grad_EL}
\end{equation}
both of which diverge as $D\to 0$.
Even in the marginal case where $\mathrm{Var}[E_L]$ is finite,
the \emph{gradient variance} (which involves these higher-order poles)
generically diverges, making gradient-based optimisation impossible at
points near the learned node.


\section{The catch-22}
\label{app:catch22:statement}

The preceding analysis reveals a three-sided dilemma:

\begin{enumerate}
\item \textbf{Backflow learns at the node.}
  The purpose of backflow is to reshape the nodal surface of the Slater
  determinant.
  The gradient signal $\partial E_L/\partial\theta_{\mathrm{BF}}$
  is largest where the backflow shifts actually change which side of the
  node a configuration point lies on, i.e., where $D\approx 0$.

\item \textbf{$E_L$ diverges at the node for imperfect $\theta$.}
  As shown in \S\ref{app:catch22:divergence}--\ref{app:catch22:mismatch},
  the local energy and its parameter gradient contain poles of the form
  $D^{-1}$ and $D^{-2}$ whenever the learned node does not coincide with
  the exact node. The resulting gradient variance is unbounded.

\item \textbf{Good collocation sampling must avoid the node.}
  Under screened collocation, the top-$K$ selection ranks points by
  $|\Psi|^2/q\propto D^2 e^{2J}/q$.
  Since $D^2\to 0$ at the node, near-node points are systematically
  excluded from the training batch. This is \emph{protective}---it
  prevents gradient explosion---but it also removes the points
  where backflow's learning signal lives.
\end{enumerate}

\noindent
A ``catch-22'' emerges:
\begin{quote}
\emph{Including near-node points gives divergent, noisy gradients and
backflow shrinks defensively or diverges.
Excluding them gives stable training but removes backflow's learning signal.
In either regime, collocation cannot effectively train backflow beyond an
accuracy floor set by the Jastrow (PINN) quality alone.}
\end{quote}


\section{Why VMC\,+\,SR escapes}
\label{app:catch22:vmc}

Under VMC, the sampling distribution \emph{is} $|\Psi_\theta|^2$, which
assigns vanishing measure to the nodal surface.
Near-node configurations are never sampled, eliminating the $D^{-1}$
singularity from the gradient estimator.
However, VMC \emph{still optimises the node position} through a
subtle mechanism:

\paragraph{Indirect nodal information through bulk correlations.}
The variational energy
$\langle E_L\rangle_{|\Psi|^2}$ is sensitive to nodal position because
shifting the node changes the wavefunction \emph{everywhere}
(through normalisation and through the functional form of $D$).
The SR force vector
$g_i=\mathrm{Cov}_{|\Psi|^2}[E_L,O_i]$
(where $O_i=\partial_{\theta_i}\log\Psi$)
receives contributions from the \emph{bulk}---regions where both
$|\Psi|^2$ and $E_L$ are well-behaved---that are
\emph{correlated with} nodal position but do not require evaluating $E_L$
at the node.

\paragraph{SR amplifies the signal.}
The Fisher metric $S_{ij}=\mathrm{Cov}[O_i,O_j]$ captures how sensitive
$\log\Psi$ is to each parameter direction.
By solving $(S+\varepsilon I)\,\delta\theta=-g$, SR
amplifies exactly those subtle bulk signals that encode nodal information,
compensating for the fact that the node itself has zero weight under
$|\Psi|^2$.

\paragraph{Core asymmetry.}
VMC obtains nodal information \emph{indirectly} through a well-conditioned
integral (the covariance $g_i$ is finite because $E_L$ is sampled away
from the node); collocation tries to obtain it \emph{directly} through
a divergent pointwise quantity ($E_L$ evaluated near the node).
This asymmetry is fundamental and cannot be resolved by
sampling tricks, loss modifications, or gradient preconditioning
within the collocation framework.


\section{Experimental evidence}
\label{app:catch22:experiments}

We present a systematic study for $N{=}6$ electrons in a 2D harmonic trap
with $\omega{=}0.5$ (DMC reference: $E_{\mathrm{DMC}}=11.78484(6)$
\cite{Pederiva2000-QD-DMC}).
The ans\"atz uses a PINN Jastrow (${\sim}15{,}000$ parameters)
and a CTNN backflow network (${\sim}8{,}600$ parameters) with
$\mathrm{bf\_scale}=0.7$.
All energies are evaluated via MCMC with 15\,000 samples after
burn-in.

\subsection{Baseline: PINN alone vs.\ PINN\,+\,CTNN}
\label{app:catch22:baseline}

\begin{table}[htbp]
\centering
\caption{Baseline collocation results for $N{=}6$, $\omega{=}0.5$.
``PINN only'' trains without backflow; ``PINN\,+\,CTNN joint'' trains both
networks from scratch for 300 epochs with $\mathrm{bf\_scale}=0.7$.
The backflow provides only a marginal improvement of ${\sim}0.1$ percentage
points.}
\label{tab:c22:baseline}
\begin{tabular}{l c c c}
\toprule
Configuration & $E$ (Ha) & $\sigma_E$ & err (\%) \\
\midrule
PINN only (300 ep) & 11.834 & 0.003 & 0.42 \\
PINN\,+\,CTNN joint (300 ep) & 11.823 & 0.003 & 0.33 \\
PINN\,+\,CTNN separate opt & 11.822 & 0.003 & 0.32 \\
\midrule
DMC reference & 11.78484 & --- & 0.00 \\
\bottomrule
\end{tabular}
\end{table}

The backflow reduces the error from $0.42\%$ to $0.32$--$0.33\%$, far short
of the $< 0.01\%$ accuracy achievable with VMC\,+\,SR on the same ansatz
(Table~\ref{tab:energies}).

\subsection{Sampling strategies}
\label{app:catch22:sampling}

We tested multiple strategies to bring near-node information into the
collocation batch, hoping to give the backflow a stronger learning signal.

\begin{table}[htbp]
\centering
\caption{Effect of sampling strategy on collocation accuracy.
All experiments use 50-epoch fine-tuning from the best
collocation checkpoint ($0.35\%$), except ``hard-smooth (scratch)''
which trains 300 epochs from random initialisation.}
\label{tab:c22:sampling}
\begin{tabular}{l l c c}
\toprule
Method & Description & err (\%) & Outcome \\
\midrule
Top-$K$ baseline & standard screened collocation & 0.35 & stable \\
Hard mining + Huber & top-$K$ of residuals & 0.41 & neutral \\
Hard mining + smoothness & + $\|\nabla^2\Delta x\|^2$ penalty & 0.38 & slight help \\
SIR (single proposal) & $|\Psi|^2$-reweighted resampling & 0.47 & worse \\
SIR (mixture) & Gaussian mixture proposal & 0.44 & worse \\
Gumbel top-$K$ & stochastic top-$K$ selection & 5.00 & catastrophic \\
\bottomrule
\end{tabular}
\end{table}

Every strategy that includes more near-node points
(SIR, Gumbel) produced \emph{worse} results---consistent with the
catch-22: the additional points carry divergent $E_L$ values that
corrupt the gradient.
Gumbel top-$K$ (which replaces deterministic selection with stochastic
Gumbel-softmax) was catastrophic, with early stopping triggered by
rapid energy divergence.

\subsection{Loss modifications}
\label{app:catch22:loss}

\begin{table}[htbp]
\centering
\caption{Effect of loss function modifications.
The hybrid loss adds a mean-energy (variational) term;
the det filter discards points with small $|\det\tilde S|$.}
\label{tab:c22:loss}
\begin{tabular}{l l c c}
\toprule
Method & Description & err (\%) & Outcome \\
\midrule
Variance loss (baseline) & Huber on $E_L-\bar E$ & 0.33 & best collocation \\
Hybrid anneal & $\beta\cdot\text{Var}+(1{-}\beta)\cdot\text{Mean}$,
  $\beta\!:\!0.7{\to}0.15$ & 2.59 & catastrophic \\
Gumbel + hybrid & stochastic top-$K$ + hybrid & 5.00 & catastrophic \\
Hybrid + det filter & $\beta{=}0.2$ fixed + $|\det|$ filter & 24.75 & broken \\
\bottomrule
\end{tabular}
\end{table}

The mean-energy gradient $\nabla_\theta\langle E_L\rangle_{\mathcal S}$
under biased collocation sampling does \emph{not} equal the VMC gradient---the
sampling distribution is not $|\Psi|^2$---so the variational principle
does not hold, and the backflow drifts to large,
unphysical shifts ($|\Delta\mathbf x|\approx 0.83$, versus ${\sim}0.23$
under variance-only training).
The variance loss's implicit constraint on backflow magnitude is
\emph{protective}, not pathological.

\subsection{Gradient regularisation}
\label{app:catch22:regularisation}

We attacked the divergence sources in $E_L$ directly:

\begin{table}[htbp]
\centering
\caption{Regularisation of the $E_L$ singularity (50-epoch
fine-tune from $0.35\%$ checkpoint). Each modification smooths or clips
one of the divergent contributions to $E_L$.}
\label{tab:c22:reg}
\begin{tabular}{l l c c}
\toprule
Method & Modification & err (\%) & Outcome \\
\midrule
Soft Coulomb & $1/r\to 1/\sqrt{r^2+\varepsilon^2}$,
  $\varepsilon{=}0.2$ & 1.41 & much worse \\
Regularised det & $\log|D|\to\operatorname{logaddexp}(\log|D|,\,
  \log\delta)$ & 0.49 & worse \\
$T$ clipping & $\max(|T|,T_{\max})$, $T_{\max}{=}30$ & 0.43 & neutral \\
\bottomrule
\end{tabular}
\end{table}

Soft Coulomb trains on wrong physics ($V$ is no longer the true potential).
Regularising the determinant distorts the $\log\Psi$ landscape.
Kinetic energy clipping is neutral because the clipped points are exactly
those that carried the (noisy) backflow signal.
None of the regularisations improve upon the baseline.

\subsection{Gradient preconditioning}
\label{app:catch22:precondition}

\begin{table}[htbp]
\centering
\caption{Gradient preconditioning experiments (300-epoch scratch training
unless noted). K-FAC uses block-diagonal Kronecker-factored Fisher on
the backflow network only.}
\label{tab:c22:kfac}
\begin{tabular}{l c c}
\toprule
Method & err (\%) & Outcome \\
\midrule
Vanilla Adam & 0.33 & baseline \\
K-FAC on CTNN & 0.34 & no improvement \\
K-FAC + smoothness & 0.51 & worse \\
Variance-minimising pretrain & 0.54 & worse \\
Adam fine-tune (30 ep, lr$=$1e-4) & 0.34 & no improvement \\
Adam burst (15 ep, lr$=$3e-4) & 0.56 & worse \\
\bottomrule
\end{tabular}
\end{table}

K-FAC provides a block-diagonal approximation to the Fisher matrix
and should, in principle, help with the ill-conditioned parameter-space
geometry.
In practice, it does not: the conditioning problem is
\emph{operator-level} (Appendices~\ref{app:laplacian}--\ref{app:coulomb}),
not merely parameter-space anisotropy, and K-FAC cannot address the
fundamental divergence in $E_L$.

\subsection{Summary of all interventions}
\label{app:catch22:summary}

\begin{table}[htbp]
\centering
\caption{Complete summary of all collocation-based backflow experiments
for $N{=}6$, $\omega{=}0.5$.
No intervention breaks the ${\sim}0.32\%$ accuracy floor.
For comparison, the same PINN\,+\,CTNN ansatz achieves $< 0.01\%$ under
VMC\,+\,SR (Table~\ref{tab:energies}).}
\label{tab:c22:summary}
\begin{tabular}{l c l}
\toprule
Experiment & err (\%) & Category \\
\midrule
PINN\,+\,CTNN sep opt (best) & 0.32 & baseline \\
PINN\,+\,CTNN joint 300 ep & 0.33 & baseline \\
K-FAC on CTNN & 0.34 & preconditioner \\
Adam fine-tune 30 ep & 0.34 & optimiser \\
Top-$K$ baseline (retrained) & 0.35 & sampling \\
Hard mining + smoothness & 0.38 & sampling \\
Hard mining + Huber & 0.41 & sampling \\
PINN only & 0.42 & no backflow \\
$T$ clipping & 0.43 & regularisation \\
SIR (mixture) & 0.44 & sampling \\
SIR (single) & 0.47 & sampling \\
Regularised det & 0.49 & regularisation \\
K-FAC + smoothness & 0.51 & preconditioner \\
Var-min pretrain & 0.54 & pretrain \\
Adam burst lr$=$3e-4 & 0.56 & optimiser \\
Soft Coulomb & 1.41 & regularisation \\
Hybrid loss (annealed) & 2.59 & loss \\
Gumbel top-$K$ + hybrid & 5.00 & sampling + loss \\
Hybrid + det filter & 24.75 & loss + filter \\
\midrule
\textbf{Same ansatz, VMC\,+\,SR} & $<$\textbf{0.01} &
  \textbf{variational} \\
\bottomrule
\end{tabular}
\end{table}

The pattern is striking:
\begin{itemize}
\item Every intervention that \emph{includes} more near-node points
  (SIR, Gumbel) \emph{increases} the error---backflow shrinks
  defensively under gradient noise or diverges entirely.
\item Every intervention that \emph{changes} the loss to tolerate
  near-node contributions (hybrid mean-energy, det filter) produces
  catastrophic results because the collocation sampling is not $|\Psi|^2$
  and the variational principle does not apply.
\item The smoothness penalty $\lambda\|\nabla^2\Delta\mathbf x\|^2$
  helps modestly (from $0.59\%$ to $0.38\%$ without it vs.\ with it)
  by damping the $\nabla^2 D/D$ poles, but cannot eliminate the
  fundamental divergence.
\item Ironically, the only method that trains backflow effectively is
  MCMC-based VMC\,+\,SR---which sidesteps the entire collocation
  framework.
\end{itemize}


\section{Connection to Appendices~\ref{app:laplacian} and \ref{app:coulomb}}
\label{app:catch22:connection}

The catch-22 is a specific manifestation of the general conditioning
difficulties analysed in Appendices~\ref{app:laplacian}--\ref{app:coulomb}:

\paragraph{Laplacian stiffness (Appendix~\ref{app:laplacian}).}
The $k^4$-type amplification of curvature errors
(\S\ref{app:laplacian:bilap}) is compounded by backflow, because
$\nabla^2(\det\tilde S)$ introduces second derivatives of
$\Delta\mathbf x$ that are not present in a fixed-node ansatz.
These additional high-frequency modes inflate the spectral spread
of the training operator $L^\ast L$.

\paragraph{Coulomb singularity (Appendix~\ref{app:coulomb}).}
The $1/r$ Coulomb potential (\S\ref{app:coulomb:setup}) adds a
singular multiplication operator to $\hat H$.
Near both the coalescence point ($r_{ij}\to 0$) and the node
($D\to 0$), $E_L$ contains products of these two singularities,
and the cusp cancellation (\S\ref{app:coulomb:cusp}) must be
maintained simultaneously with the nodal cancellation
\eqref{eq:c22:exact_cancel}.
In the strong residual squared $\|LE_L\|^2$, these compound
singularities create an extremely stiff landscape.

\paragraph{The backflow-specific pathology.}
What distinguishes the backflow catch-22 from the general conditioning
problem is the \emph{self-referential} nature of the singularity:
the backflow parameters $\theta_{\mathrm{BF}}$ determine \emph{where}
the node is, so updating $\theta_{\mathrm{BF}}$ moves the singularity
in configuration space.
Under the collocation loss, the gradient pushes $\theta_{\mathrm{BF}}$
in a direction determined by $E_L$ values at the current collocation
points, but the resulting node shift invalidates those very points.
VMC+SR avoids this because it samples from the \emph{current}
$|\Psi_\theta|^2$ at every step, so the sampling distribution
co-adapts with the node position.

